---
title: "Basic Statistics by Wilcox"
author: "Granitz, Stephan"
date: "February 3, 2018"
output: html_document
---

---  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(ggalt)
```

##### TOC
.                               .  | .                                    .
---------------------------------- | ---------------------------------------
1 - Numerical Summaries of Content | 7 - Correlation & Regression
2 - Graphical Summaries of Content | 8 - Comparing Two Groups
3 - Probability                    | 9 - Comparing More Than Two Groups
4 - Sampling Distributions         | 10- Multiple Comparisons
5 - Estimation                     | 11- Categorical Data
6 - Hypothesis Testing             | 12- Rank Based & Nonparametric Methods
  
#### USEFUL SYMBOLS
.                               .  | .                                    .
---------------------------------- | ---------------------------------------
$\alpha$ - _alpha_ - probability of Type I error | $\nu$ - _nu_ - degrees of freedom
$\beta$ - _beta_ - probability of Type II error | $\Omega$ - _Omega_ - odds ratio
$\beta$~1~ - slope of regression line | $\rho$ - _rho_ - population correlation coefficient
$\beta$~0~ - intercept of regression line | $\sigma$ - _sigma_ - population standard deviation
$\delta$ - _delta_ - measure of effect size | $\phi$ - _phi_ - measure of association
$\epsilon$ - _epsilon_ - residual / error term in ANOVA regression | $\chi$ - _chi_ - $\chi$^2^ chi-squared distribution
$\theta$ - _theta_ - odds ratio or population median | $\Delta$ - _Delta_ - measure of change / effect size
$\mu$ - _mu_ - population mean | $\Sigma$ - _Sigma_ - summation
$\mu$~t~ - trimmed population mean | $\tau$ - _tau_ - Kendall's tau


> Modern statistics provides a quantitative technology for empirical sciences. It is a logic and methodology for the measurement of uncertainty and for examination of the consequences of that uncertainty in the planning and interpretation of experimentation and observation  

#### IMPORTANT WORKS
.                           .  | .                                .
------------------------------ | -----------------------------------
1749-1827 Laplace, Simon | 1894-1981 Neyman, Jerzy
1777-1855 Gauss, Carl | 1895-1980 Pearson, Egon
1857-1936 Pearson, Karl | 1890-1962 Fisher, Ronald

#### THREE FUNDAMENTAL COMPONENTS
1. *Desgin* - the planning and carrying out of a study
2. *Description* - the methods for summarizing data
3. *Inference* - making predictions or generalizations about a population of individuals or things based on a sample of observations

### 1. Numerical Summaries of Data
"Descriptive Statistics"  
  
  
Useful Symbols                     | Summation Notation
---------------------------------- | ---------------------------------------
$X$ $\rightarrow$ represents a random variable | $\sum_{i=1}^{n}X_{i}=X_{1}+X_{2}+...+X_{n}$
$c$ $\rightarrow$ constant | sum all $X$ from $X_{i}$ to $X_{n}$
$i$ $\rightarrow$ index | $\sum cX_{i}= c\sum X_{i}$
$n$ $\rightarrow$ total number of observations | sum of all variables times a constant equals the constant times the sum of all variables
  
**Measures of location** (_measures of central tendency_) - single numerical quantity that reflects _typical_ observed value  
  
  Mean $\rightarrow$ $\overline{X} = \frac{1}{n}\sum x_{i}$  
  _The mean can be sensitive to even one outlier_  
    
  Median $= \left\{\begin{matrix}
n \rightarrow odd & M = X_{m} & m = \frac{(n+1)}{2} \\ 
n \rightarrow even & M = \frac{(X_{m} + X_{m+1})}{2} & m = \frac{n}{2}
\end{matrix}\right.$  
  _The median is a **resistant** measure of location because it is insensitive to outliers_
   
**Measures of dispersion** (_measures of variation_) - reflects how spread out the data happens to be  
  
  Range $\rightarrow$ $X_{max}-X_{min}$  
  
  Variation $\rightarrow$ $s^{2} = \frac{1}{n-1}\sum (X_{i} - \overline{X})^{2}$  
  
  Standard Deviation $\rightarrow$ $s = \sqrt[]{s^{2}}$  
   
**Five number summary of data**  
1. Lowest observed value $X_{min}$  
2. $Q_{1}$ = lower quartile (median of bottom half)  
3. Median (middle value, cuts data in half)  
4. $Q_{3}$ = upper quartile (median of upper half)  
5. Largest observed value $X_{max}$  
   
$IQR$ $\rightarrow$ interquartile range $= Q_{3} - Q_{1}$  
    
Outliers $= \left\{\begin{matrix}
X < Q_{1} - IQR\\ 
X > Q_{3} + IQR
\end{matrix}\right.$  
  
  
```{r numerical_summaries, echo=FALSE}
# Example of Numerical Summaries
n <- 300
X <- rnorm(n, mean = 0, sd = 1)

x_bar <- mean(X)
five_num <- fivenum(X)
label <- c("Min", "Q1", "Median", "Q3", "Max")
med <- five_num[3]

ggplot() + 
  geom_point(aes(X, rep(3, n)), color="lightgrey", alpha=0.33, position = "jitter") +
  geom_point(aes(five_num, rep(3, 5)), color="indianred", size=3) +
  annotate("text", x = five_num, y = 2.75, label = label, color = "indianred", fontface = 2) +
  geom_point(aes(x_bar, 3), color="purple", size=3, alpha=0.5) +
  annotate("text", x = x_bar, y = 3.25, label = "Mean", color = "purple", fontface = 2) +
  geom_line(aes(c(x_bar - sd(X), x_bar), 3.5), color = "blue") +
  annotate("text", x = x_bar - sd(X) / 2, y = 3.65, label = "Standard Deviation", color = "blue") +
  geom_line(aes(c(five_num[c(2,4)]), 2.5), color = "darkgreen") +
  annotate("text", x = med, y = 2.4, label = "IQR", color = "darkgreen") +
  geom_line(aes(c(five_num[c(1,5)]), 2.25), color = "purple") +
  annotate("text", x = med, y = 2.1, label = "Range", color = "purple") +
  ylim(1,4) + theme_bw() + labs(title = "Example of Numerical Summaries", x = "X", y = "") +
  theme(axis.title.y = element_blank(),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank())


```


### 2. Graphical Summaries of Data

Notation                           | Meaning
---------------------------------- | ---------------------------------------
$fx$ | frequency, "value of $x$ occured $fx$ times"
$max(fx)$ | mode of $x$
$fx/n$ | relative frequency, plot of relative frequencies is called _empirical distribution_
$\Sigma fx = n \rightarrow \frac{1}{n} \Sigma fx = 1$ | sum of relative frequencies is always 1
$\bar{x} = \frac{1}{n} \Sigma xfx = \Sigma x\frac{fx}{n}$ | sample mean
$s^2 = \frac{n}{n-1} \Sigma \frac{fx}{n}(x - \bar{X})^2$ | sample variance

_Cumulative relative frequency distribution_ is the proportion of observations less than or equal to a given value  
- for every $x$, some proportion of observed values will be $\leqslant x$ (denoted by $F(x)$)  
- these proportions taken as a whole are the _empirical cumulative distribution function_  

```{r rel_freq, echo=FALSE}
freq <- data.frame(x=0:5, y=c(0.1, 0.2, 0.15, 0.39, 0.12, 0.04))

as_tibble(freq) %>% print()

ggplot(freq, aes(x, y)) + 
  geom_lollipop(point.colour="steelblue", point.size=3) +
  geom_line(aes(x, cumsum(y)), color="lightgrey") +
  annotate(
    "text", 
    x = freq$x, 
    y = cumsum(freq$y) + 0.05, 
    label = as.character(cumsum(freq$y)), 
    color = "grey"
  ) +
  annotate(
    "text", 
    x = max(freq$x) / 2 + 1, 
    y = 0.7, 
    label = "Cumulative relative \n frequency distribution", 
    color = "grey"
  ) +
  annotate(
    "text", 
    x = freq$x, 
    y = freq$y + 0.05, 
    label = as.character(freq$y), 
    color = "darkgreen"
  ) +
  annotate(
    "text", 
    x = freq$x[freq$y == max(freq$y)] + 0.25, 
    y = max(freq$y), 
    label = "mode", 
    color = "purple"
  ) +
  theme_minimal()
```

_Histogram_ is similar to the relative frequency plot with the addition of *bins*  
Bins group data into defined categories and then plot the corresponding frequencies  

```{r hist, echo=FALSE}
hist <- tibble(cat = c("0-0.5","0.5-1","1-1.5","1.5-2","2-2.5","2.5-3"), 
               freq = c(10,20,18,12,3,2)) %>% 
        mutate(rel_freq = round(freq/sum(freq), 2)) 

print(hist)

hist %>%
  separate(cat, into = c("min", "max"), sep = "-") %>% 
  mutate(min = as.numeric(min), max = as.numeric(max)) %>% 
  ggplot() + 
    geom_segment(aes(x = min, xend = min, y = 0, yend = freq)) +
    geom_segment(aes(x = max, xend = max, y = 0, yend = freq)) +
    geom_segment(aes(x = min, xend = max, y = 0, yend = 0)) +
    geom_segment(aes(x = min, xend = max, y = freq, yend = freq))
```

